%%% Discussion

More broadly, implicit pangenome graphs introduce a new layer in the pangenome analysis stack. Rather than requiring a single monolithic construction step, the alignment-space framework lets users query, partition, and extract regions on demand, then selectively invoke graph construction only where explicit topology is needed---for bubble enumeration, snarl decomposition, or graph-aware variant calling via tools like Minigraph-Cactus \cite{Hickey2023} or personalized pangenome references \cite{Siren2024}. IMPG can supply these builders with partitioned input through its partition-lace pipeline, or materialize graphs on demand for specific loci. The two representations are complementary: alignment space provides the upstream infrastructure for exploration and partitioning, while graph space provides the downstream topology for variant-aware analyses.

IMPG's projection accuracy is determined by the quality of the input alignments: given correct alignments, exact-mode queries---which traverse full CIGAR strings (PAF) or reconstruct them via wavefront alignment from tracepoints (.1aln/.tpa)---produce base-exact coordinates by construction. For tracepoint-based formats, an approximate query mode bypasses CIGAR reconstruction entirely, estimating projected boundaries from tracepoint segment coordinates; this is faster but boundaries may differ from exact projections by up to the tracepoint segment size per hop (e.g., $\leq$100\,bp for fixed-spacing tracepoints; \textcolor{red}{Supplementary Table~X}). Regions absent from input alignments cannot be recovered. The incremental design means each new assembly requires alignment against only a spanning subset of existing haplotypes, not a complete rebuild---enabling continuous growth of the pangenome index. Looking forward, transitive queries yield pairwise similarity across indexed haplotypes, opening the possibility of computing population genetics summary statistics directly from alignment space without decomposing variation into biallelic sites. By operating in alignment space, IMPG makes it possible to study sequence variation across populations at scale without requiring upfront graph construction---bridging the gap between assembly production and population-level genomic analysis.
